---
title: "Approaches based on CIF specification"
author: "Karla Monterrubio-Gómez, Nathan Constantine-Cooke, and Catalina Vallejos"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: show
    toc: yes
    css: style.css
    theme: simplex
    highlight: textmate
    toc_float:
      collapsed: no
link-citations: yes
citation_package: natbib
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}

---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


# Dataset

In order to demonstrate the methods, we employ publicly available data. 

The dataset used here corresponds to the Hodgkin’s disease (HD) study described
in Pintilie, 2006. The dataset comprises 865 patients diagnosed with early stage
(I or II) HD, and which were treated either with radiation (RT) or with
radiation and chemotherapy (CMT).

The recorded data includes:

* age: Age (years)
* sex: Sex, F=female and M=Male.
* trtgiven: Treatment given, RT=Radiation, CMT=Chemotherapy and radiation
* medwidsi: Size of mediastinum involvement, N=No, S=Small, L=Large
* extranod: Extranodal disease, Y=Extranodal disease, N= Nodal disease
* clinstg: Clinical stage, 1=Stage I, 2=Stage II
* time: time to failure (years) calculated from the date of diagnosis
* status: 0=censoring, 1=relapse and 2=death.

We now load and display the structure of the HD dataset:

```{r, message=FALSE, warning=FALSE}
library(readr)
hd <- data.frame(read_csv("../Data/HD/hd.csv",
                          col_types = cols(X1 = col_skip())))
str(hd)
```

To proceed with the analysis, it is important to change the data type of sex,
trtgiven, medwidsi, and extranod from character to factor. Similarly, we convert
clinstg from numeric to factor.

```{r}
hd$sex      <- as.factor(hd$sex)
hd$trtgiven <- as.factor(hd$trtgiven)
hd$medwidsi <- as.factor(hd$medwidsi)
hd$extranod <- as.factor(hd$extranod)
hd$clinstg  <- as.factor(hd$clinstg)
```

Now, we explore the number of events for each event type:

```{r}
table(hd$status)
```

Thus, we have 439 censored patients, 291 with relapse, and 135 who died. From
now on, we assume that the event of interest is relapse, i.e. status=1.

In order to create a test set, we use stratified sampling to partition our
dataset into 80% for train and 20% for test.

```{r, eval=FALSE}
library(splitstackshape)
split_data <- stratified(hd, c("status"), 0.8, bothSets = TRUE)
hd_train   <- split_data$SAMP1
hd_test    <- split_data$SAMP2
```


```{r, echo=FALSE}
hd_train <- read_csv("../Data/HD/hd_train.csv", show_col_types = FALSE)[, -1]
hd_test <- read_csv("../Data/HD/hd_test.csv", show_col_types = FALSE)[, -1]

hd_train$sex      <- as.factor(hd_train$sex)
hd_train$trtgiven <- as.factor(hd_train$trtgiven)
hd_train$medwidsi <- as.factor(hd_train$medwidsi)
hd_train$extranod <- as.factor(hd_train$extranod)
hd_train$clinstg  <- as.factor(hd_train$clinstg)

hd_test$sex      <- as.factor(hd_test$sex)
hd_test$trtgiven <- as.factor(hd_test$trtgiven)
hd_test$medwidsi <- as.factor(hd_test$medwidsi)
hd_test$extranod <- as.factor(hd_test$extranod)
hd_test$clinstg  <- as.factor(hd_test$clinstg)
```

Now, we explore the number of observations per status in both train and test
set:

```{r}
table(hd_train$status)
table(hd_test$status)
```

# Fine-Gray 

In contrast to models that use a cause-specific CPH specification, this approach
permits to do inferences on the covariate effects on both, the hazard function
and the survival function due to the fact that competing events are treated
differently.

As with cause-specific CPH models, there are several R packages to fit a
sub-distribution hazard model. Here we explore `cmprsk` and `riskRegression`.

## `cmprsk`

In order to fit the model, one should first create a design matrix containing
the covariates of interest, this can be done employing the `model.matrix()`
function, which creates dummy variables for the discrete predictors.

```{r, message=FALSE}
library(cmprsk)
predictors <- model.matrix(~ age +
                             sex +
                             trtgiven +
                             medwidsi +
                             extranod +
                             clinstg,
                           data = hd_train)
head(predictors)
```

Note that our covariates now corresponds to indicators (1 or 0) of a specific
level of the original factor variables. In order to employ this new matrix, we
will first drop the intercept. Once that has been done, we can employ the
`crr()` function to fit the model.

```{r}
predictors <-  predictors[, -1] # drop first column (intercept)


FG_crr <- crr(ftime = hd_train$time,
              fstatus = hd_train$status,
              cov1 = predictors, # for time interactions use argument cov2
              failcode = 1,
              cencode = 0)
summary(FG_crr)
```
The above output includes the estimated covariate effects ($\exp(\beta)$), as
well as significance level from where we can infer that age, treatment given,
and clinical stage are significant. Note however that estimated HR are not
identical to those inferred in the cause-specific vignette. 

In this setting, we can make inferences on both, the covariate effects on the
sub-distribution hazard and the covariate effects on the CIF. For instance, the
sub-distribution hazard ratio of relapse is 1.363 when a patient is in clinical
stage 2. In addition, because the hazard is greater that one, one can also infer
that being in clinical stage 2 increases the incidence of relapse. Here is
important to highlight that we can not use the magnitude of the hazard ratios to
make inferences on the magnitude of the effect on the probability of occurrence.

In order to check the proportional hazard assumption we use Schoenfeld residuals
plotted against time for each of the covariates. If the proportional assumption
is violated, one expects to see a non-constant mean across time.

```{r, figures-side, fig.show="hold", out.width="47%"}
par(mar = c(4, 4, 2, 0.1))
for (j in seq_len(ncol(FG_crr$res))){
scatter.smooth(FG_crr$uftime,
               FG_crr$res[, j],
               main = colnames(predictors)[j],
               xlab = "time",
               ylab = "Schoenfeld residuals")
}
```

In order to do predictions, we also convert our test data (hd_test) into a
design matrix format. Then, the function `predict()` will return the estimated
CIF at all unique time events for all patients in the test set. This function
computes $1-\exp(-H)$, where $H$ is the cumulative hazard obtained using the
Breslow estimator.

```{r}
predictors_test <- model.matrix(~age +
                                  sex +
                                  trtgiven +
                                  medwidsi +
                                  extranod +
                                  clinstg,
                                data = hd_test)[, -1]

CIF_hat <- predict(object = FG_crr, cov1 = predictors_test)

dim(CIF_hat)
```

The first column in CIF_hat corresponds to the 68 unique event times and the
remaining 173 columns are the estimated CIFs for each patient at such event
times. For instance, the cumulative incidence function of patient 1 (63 years,
male, under radiation, no mediastinum involvement, no extranodal disease, and
clinical stage 1) and patient 3 (27 years, female, no treatment radiation,
mediastinum involvement, no extranodal disease, and clinical stage 2) are shown
below in red and black, respectively.

```{r, fig.dim=c(6,4)}
par(mar = c(4, 4, 2, 0.1))
plot(CIF_hat[, 1],
     CIF_hat[, 2],
     type = "l",
     col = "red",
     ylim = c(0, 0.5),
     xlab = "Time (years)",
     ylab = "Cumulative incidence")

lines(CIF_hat[, 1], CIF_hat[, 4])
```

**Can we do predictions at a specific time point using cmprsk? - doesn't look like**

**ADD HERE model performance**

## `riskRegression`

This R package acts as a wrapper of the `crr()` function described above.
However, it permits to directly specify the model through the formula argument,
where we use `Hist()` to obtain the the response variable. When the cause
argument is not passed the default is set to 1.

Below we confirm that estimated coefficients are identical to those obtained
with `cmprsk` package and the interpretation is the same.

```{r, message=FALSE}
library(prodlim)         # for Hist()
library(riskRegression)

FG_riskregression <- FGR(formula = Hist(time, status) ~
                           age +
                           sex +
                           trtgiven +
                           medwidsi +
                           extranod +
                           clinstg,
                          data = hd_train,
                          cause = 1)
print(FG_riskregression)
```
**how to get residuals here?**

Using this package we can obtain predictions of the CIF at a specific time point
(e.g. $t=30.5$) for a new dataset (hd_test). Below, we show results for the
first 5 subjects in the test set.

```{r}
FG.prediction <- predict(FG_riskregression, newdata = hd_test,  times = 30.5)
FG.prediction[c(1:5)]
```

Because the Fine-Gray model can be recovered when a complementary log-log link
function is used in a transformation model (Gerds et.al 2012),
**change references** we can use `riskRegression()` with `link = "prop"` as
shown below

```{r}
FG.prop <- riskRegression(formula = Hist(time, status) ~
                            age +
                            sex +
                            trtgiven +
                            medwidsi +
                            extranod +
                            clinstg,
                          data = hd_train,
                          cause = 1,
                          link = "prop")
summary(FG.prop)
```

In this case, `predict()` returns the risk at observed time points for each
subject.

```{r}
prediction.FG.prop  <- predict(FG.prop, newdata = hd_test[c(1, 3), ])
prediction.FG.prop
```

Below, we plot the CIF for patient 1 (red) and 3 (black) in the test set.

```{r, fig.dim=c(6,4)}
par(mar = c(4, 4, 2, 0.1))
plot(prediction.FG.prop$time,
     prediction.FG.prop$risk[1, ],
     type = "l",
     col = "red",
     ylim = c(0, 0.5),
     xlab = "Time (years)",
     ylab = "Cumulative incidence")
lines(prediction.FG.prop$time, prediction.FG.prop$risk[2, ])
```


# Sparse regression

**CoxBoost was deleted from CRAN**

# Pseudo-values

**Note that the package have been updated with respect to the code shown in Klein, 2008. Thus, the user is referred to the help of the function.**

We define a time point grid of $M=7$ equally spaced points at which we will
compute the pseudo-values. Pseudo-observations are computed with the
`pseudoci()` function for both competing events. Below, we only consider
relapse (cause 1).

```{r, message=FALSE}
library(pseudo)

cutoffs <- c(5, 10, 15, 20, 25, 30, 35)

pseudo <- pseudoci(hd_train$time,
                   hd_train$status,
                   tmax = cutoffs)$pseudo$cause1     # we only consider cause 1
dim(pseudo)
```

pseudo contains the pseudo-values of the 692 patients (rows) at the 7 different
time points (columns) for relapse. Below, we show the estimated CIF for patient
109 (45 years old, male, under radiation, and stage 1) who had relapse at time
6.

```{r, fig.dim=c(6,4)}
plot(pseudo[109, ], type = "l", xlab = "Time", ylab = "Pseudo-value")
```

The following code, will merge the pseudo-values with our train dataset.

```{r}
merged_dataset <- NULL
for (j in seq_len(ncol(pseudo))) {
    merged_dataset <- rbind(merged_dataset,
                            cbind(hd_train,
                                  pseudo = pseudo[, j],
                                  tpseudo = cutoffs[j],
                                  id = seq_len(nrow(hd_train))))
}

merged_dataset <- merged_dataset[order(merged_dataset$id), ]
merged_dataset$tpseudo <- as.factor(merged_dataset$tpseudo)
```

We will now fit a generalised linear model using `geese()` function from R
package `geepack`. We set `mean.link = "cloglog"` to use a complementary
log-log link in order to recover Fine-Gray approach. 

```{r, message=FALSE}
library(geepack)

fit_pseudo <- geese(pseudo ~
                      tpseudo +
                      age +
                      sex +
                      trtgiven +
                      medwidsi +
                      extranod +
                      clinstg - 1,
                    data = merged_dataset,
                    id = id,
                    jack = TRUE,
                    scale.fix = TRUE,
                    family = gaussian,
                    mean.link = "cloglog",
                    corstr = "independence")
summary(fit_pseudo)
```

The reported coefficients for the selected time points are used to estimate the
CIF. 

Covariate effects are shown below and are interpreted as logarithms of
sub-distribution hazard ratios. We observe that the estimated effects are akin
to those obtained in the previous section.

```{r}
exp(fit_pseudo$beta)
```

**How to do predictions at a new time point here?**

# Direct binomial

Binomial modelling approach can be fitted using the `timereg` R package.

To fit the model, we should first use the `Event()` function to obtain the
response variable to be used in the formula argument of the `comp.risk()`
function. The cause of interest must be explicitly defined through the cause
argument. The function supports several models. Here we focus on recovering the
Fine Gray approach.

```{r, warning=FALSE}
library(timereg)

fit_DB <- comp.risk(Event(time, status) ~
                      const(age) +
                      const(sex) +
                      const(trtgiven) + # const for time-invariant
                      const(medwidsi) +
                      const(extranod) +
                      const(clinstg),
                    data = hd_train,
                    cause = 1,
                    n.sim = 5000,
                    model = "prop",       # Fine-Gray
                    resample.iid = 1,
                    # uses all covariates present for the censoring weights
                    cens.model = "cox")
summary(fit_DB)
```

According to the summary output;  age and treatment are significant. Below, we
show estimated hazard ratios.

**NCC this is broken**
```{r, eval = FALSE}
knitr::kable(round(cbind(exp(fit_DB$gamma), 4)), col.names = "Exp coef")
```


In order to do predictions of the CIF, we use the `predict()` function. The
default makes predictions at all observed time points, but we can also select a
specific time through the argument times.

```{r fig.dim=c(6,4)}
par(mar = c(4, 4, 2, 0.1))
hd_test <- as.data.frame(hd_test[, 1:6])
pred_DB <- predict.timereg(fit_DB, newdata = hd_test[c(1, 3), 1:6], se = 0)
plot(pred_DB,
     multiple = 1,
     col = c("red", "black"),
     ylim = c(0, 0.57),
     ylab = "Cumulative incidence")

```

Absolute risk predictions at time $t=30.5$ for the first 5 patients are shown:

```{r}
pred_DB.t <- predict.timereg(fit_DB, newdata = hd_test[1:5, 1:6], times = 30.5)
c(pred_DB.t$P1)
```

`timereg` permits fitting the model time varying effects. **warning problem**

```{r}
fit_DBtime <- comp.risk(Event(time, status) ~
                          age +
                          sex +
                          trtgiven +
                          medwidsi +
                          extranod +
                          clinstg,
                        data = hd_train,
                        cause = 1,
                        n.sim = 5000,
                        model = "prop", # Fine-Gray
                        resample.iid = 1,
                        # uses all covariates present for the censoring weights
                        cens.model = "cox")
summary(fit_DBtime)
```

# Dependent Dirichlet processes

The formula definition uses `Hist()`.

Indication of a competing risk setting must be passed by setting
`comp = TRUE`.  Below, we run the MCMC chain for 5000 iterations, we discarded
3000 as part of the burn-in period, and to reduce autocorrelation, a thinning of
10 was used (default).

**NCC Make comment about DPWeibull not being avialble on CRAN? The package is not on CRAN because a dependency, binaryLogic, was removed from CRAN as the maintainer's email address is no longer accessible**

```{r, warning=FALSE}
library(DPWeibull)
library(prodlim) # for Hist() function
fit_DDP <- dpweib(Hist(time, status) ~
                    age +
                    sex +
                    trtgiven +
                    medwidsi +
                    extranod +
                    clinstg ,
                  data = hd_train,
                  comp = TRUE,
                  burnin = 1000,
                  iteration = 5000,
                  predtime = seq(from = min(hd_train$time),
                                 to = max(hd_train$time),
                  length =  40))
summary(fit_DDP)
```

One can explore trace plots of the estimated parameters using the `coda`
library. This will help us to determine if the MCMC parameters employed when
fitting the model should be modified. For instance, below we show trace plots of
$\alpha$ and $\lambda$ for patient 1.

```{r, message=FALSE, fig.dim=c(6,4)}
library(coda)
par(mar = c(4, 4, 2, 0.1))
traceplot(as.mcmc(fit_DDP$alpharec1[, 1]))
traceplot(as.mcmc(fit_DDP$lambda0rec1[, 1]))
```

```{r}
(fit_DDP$betarec1[1:3, 1:3])
dim(fit_DDP$loghr.est)
fit_DDP$predtime
(fit_DDP$xmean)
fit_DDP$covnames
```

To do predictions,

```{r fig.dim=c(6,4)}
par(mar = c(4, 4, 2, 0.1))
hd_test_subset <- as.data.frame(hd_test[c(1, 4), 1:6])

pred_DDP <- predict(fit_DDP, newdata = hd_test_subset)
plot(pred_DDP$tpred,
     pred_DDP$Fpred[2, ],
     type = "l",
     col = "red",
     xlab = "Time (years)",
     ylab = "Cumulative incidence")
lines(pred_DDP$tpred, pred_DDP$Fpred[1, ])
plot(pred_DDP)
```
 

# References


* Gerds, T. A., Scheike, T. H., & Andersen, P. K. (2012). Absolute risk regression for competing risks: interpretation, link functions, and prediction. Statistics in medicine, 31(29), 3921-3930.
* Klein, J. P., Gerster, M., Andersen, P. K., Tarima, S., & Perme, M. P. (2008). SAS and R functions to compute pseudo-values for censored data regression. Computer methods and programs in biomedicine, 89(3), 289–300. https://doi.org/10.1016/j.cmpb.2007.11.017



# Session Info
```{r}
sessionInfo()
```
