---
title: "Approaches based on CIF specification"
author: "Karla Monterrubio-Gómez, Nathan Constantine-Cooke, and Catalina Vallejos"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: show
    toc: yes
    css: style.css
    theme: simplex
    highlight: textmate
    toc_float:
      collapsed: no
link-citations: yes
bibliography: references.bib
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}

---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


# Dataset

In order to demonstrate the methods, we employ publicly available data. 

The dataset used here corresponds to the Hodgkin’s disease (HD) study described
in @pintilie2006competing. The dataset comprises 865 patients diagnosed with early stage
(I or II) HD, and which were treated either with radiation (RT) or with
radiation and chemotherapy (CMT).

The recorded data includes:

* age: Age (years)
* sex: Sex, F=female and M=Male.
* trtgiven: Treatment given, RT=Radiation, CMT=Chemotherapy and radiation
* medwidsi: Size of mediastinum involvement, N=No, S=Small, L=Large
* extranod: Extranodal disease, Y=Extranodal disease, N= Nodal disease
* clinstg: Clinical stage, 1=Stage I, 2=Stage II
* time: time to failure (years) calculated from the date of diagnosis
* status: 0=censoring, 1=relapse and 2=death.

We now load and display the structure of the HD dataset:

```{r, message=FALSE, warning=FALSE}
library(readr)
hd <- data.frame(read_csv("../Data/HD/hd.csv",
                          col_types = cols(X1 = col_skip())))
str(hd)
```

To proceed with the analysis, it is important to change the data type of sex,
trtgiven, medwidsi, and extranod from character to factor. Similarly, we convert
clinstg from numeric to factor.

```{r}
hd$sex      <- as.factor(hd$sex)
hd$trtgiven <- as.factor(hd$trtgiven)
hd$medwidsi <- as.factor(hd$medwidsi)
hd$extranod <- as.factor(hd$extranod)
hd$clinstg  <- as.factor(hd$clinstg)
```

Now, we explore the number of events for each event type:

```{r}
table(hd$status)
```

Thus, we have 439 censored patients, 291 with relapse, and 135 who died. From
now on, we assume that the event of interest is relapse, i.e. status=1.

In order to create a test set, we use stratified sampling to partition our
dataset into 80% for train and 20% for test.

```{r, eval=FALSE}
library(splitstackshape)
set.seed(2022)
split_data <- stratified(hd, c("status"), 0.8, bothSets = TRUE)
hd_train   <- split_data$SAMP1
hd_test    <- split_data$SAMP2
```

Now, we explore the number of observations per status in both train and test
set:

```{r}
table(hd_train$status)
table(hd_test$status)
```

# Fine-Gray 

In contrast to models that use a cause-specific CPH specification, this approach
permits inferences for the covariate effects on both the hazard function
and the survival function due to the fact that competing events are treated
differently.

As with cause-specific CPH models, there are several R packages to fit a
sub-distribution hazards model. Here we explore `cmprsk` and `riskRegression`.

## `cmprsk`

In order to fit the model, one should first create a design matrix containing
the covariates of interest, this can be done employing the `model.matrix()`
function, which creates dummy variables for the discrete predictors.

```{r, message=FALSE}
library(cmprsk)
predictors <- model.matrix(~ age +
                             sex +
                             trtgiven +
                             medwidsi +
                             extranod +
                             clinstg,
                           data = hd_train)
head(predictors)
```

Note that our covariates now corresponds to indicators (1 or 0) of a specific
level of the original factor variables. In order to employ this new matrix, we
will first drop the intercept. Once that has been done, we can employ the
`crr()` function to fit the model.

```{r}
predictors <-  predictors[, -1] # drop first column (intercept)


FG_crr <- crr(ftime = hd_train$time,
              fstatus = hd_train$status,
              cov1 = predictors, # for time interactions use argument cov2
              failcode = 1,
              cencode = 0)
summary(FG_crr)
```

The above output includes the estimated covariate effects ($\exp(\beta)$), as
well as significance level from which we can infer that the marginal effects for
age, treatment given, and clinical stage are significant. Note however that 
estimated hazard ratios are non-identical to those inferred in the 
cause-specific vignette. This is because the risks set differ between these two 
hazard functions.

In this setting, we can make inferences on both, the covariate effects on the
sub-distribution hazard and the directional effects of covariates on the CIF. 
For instance, the sub-distribution hazard ratio of relapse is 
`r round(exp(FG_crr$coef["clinstg2"]),3)` when a
patient is in clinical stage 2. In addition, because the hazard ratio is greater
than one, one can also infer that being in clinical stage 2 increases the
incidence of relapse. Here, is important to highlight that we cannot use the
magnitude of the hazard ratios to make inferences on the magnitude of the effect
on the probability of occurrence.

Because the sub-distribution hazard is directly related to the cumulative
incidence function, we can make risk predictions. In order to do this, we first
convert our test data (`hd_test`) into a design matrix format. Then, the function
`predict()` will return the estimated CIF at all unique time events for all
patients in the test set. This function computes $1-\exp(-H)$, where $H$ is the
cumulative hazard obtained using the Breslow estimator.

```{r}
predictors_test <- model.matrix(~ age +
                                  sex +
                                  trtgiven +
                                  medwidsi +
                                  extranod +
                                  clinstg,
                                data = hd_test)[, -1]

CIF_hat <- predict(object = FG_crr, cov1 = predictors_test)

dim(CIF_hat)
```

The first column in CIF_hat corresponds to the 68 unique event times and the
remaining 173 columns are the estimated CIFs for each patient (given their 
covariate values) at such event times. As an illustration, the following 
code is used to plot the estimated CIF for events type 1 for two patients in the
training dataset. The first patient is 37 years old, male, under 
radiation treatment, no mediastinum involvement, no extranodal disease, and 
clinical stage1.The second patient is 41 years old, male, under radiation, 
no mediastinum involvement, no extranodal disease, and clinical stage 2. 

```{r, fig.dim=c(6,4)}
par(mar = c(4, 4, 2, 0.1))
plot(CIF_hat[, 1],
     CIF_hat[, 2],
     type = "l",
     col = "red",
     ylim = c(0, 0.5),
     xlab = "Time (years)",
     ylab = "Cumulative incidence")

lines(CIF_hat[, 1], CIF_hat[, 4], col="blue")
legend("topright", legend = c("Patient 1", "Patient 2"),
       lty = c(1,1), col = c("red", "blue"))
```

> **_NOTE:_** Using `cmprsk` it is not straightforward to do predictions at a 
specific time point beyond the unique time-points present in the training data. 

## `riskRegression`

This R package acts as a wrapper of the `crr()` function described above.
However, it permits directly specifying the model through the formula argument,
where we use `Hist()` to obtain the the response variable. When the cause
argument is not passed, the default is 1.

Below we confirm that estimated coefficients are identical to those obtained
with `cmprsk` package and the interpretation is the same.

```{r, message=FALSE}
library(prodlim)         # for Hist()
library(riskRegression)

FG_riskregression <- FGR(formula = Hist(time, status) ~
                           age +
                           sex +
                           trtgiven +
                           medwidsi +
                           extranod +
                           clinstg,
                          data = hd_train,
                          cause = 1)
print(FG_riskregression)
```

Differently to `cmprsk`,`riskRegression` permits to do predictions of the CIF at
a specific time point (e.g. $t=5$ years) for a new dataset (e.g. `hd_test`). Below, 
we show results for the first 5 subjects in the test set.

```{r}
FG.prediction <- predict(FG_riskregression, newdata = hd_test,  times = 5)
FG.prediction[c(1:5)]
```

Because the Fine-Gray model can be recovered when a complementary log-log link
function is used in a transformation model (@gerds2012), we can use
`riskRegression()` with `link = "prop"` as shown below:

```{r}
FG.prop <- riskRegression(formula = Hist(time, status) ~
                            age +
                            sex +
                            trtgiven +
                            medwidsi +
                            extranod +
                            clinstg,
                          data = hd_train,
                          cause = 1,
                          link = "prop")
summary(FG.prop)
```

In this case, `predict()` returns the risk at observed time points for each
subject.

```{r}
prediction.FG.prop  <- predict(FG.prop, newdata = hd_test[c(1, 3), ])
prediction.FG.prop
```

Below, we plot the CIF for patient 1 (red) and 2 (blue) in the test set.

```{r, fig.dim=c(6,4)}
par(mar = c(4, 4, 2, 0.1))
plot(prediction.FG.prop$time,
     prediction.FG.prop$risk[1, ],
     type = "l",
     col = "red",
     ylim = c(0, 0.5),
     xlab = "Time (years)",
     ylab = "Cumulative incidence")
lines(prediction.FG.prop$time, prediction.FG.prop$risk[2, ], col="blue")
legend("topright", legend = c("Patient 1", "Patient 2"),
       lty = c(1,1), col = c("red", "blue"))
```


# Sparse regression

**CoxBoost was deleted from CRAN**

# Pseudo-values

> **_NOTE:_** the package have been updated with respect to the code shown in
@klein2008sas. The user is referred to the package documentation for the
function.**

We define a time point grid of $M=7$ equally spaced points at which we will
compute the pseudo-values. Pseudo-observations are computed with the
`pseudoci()` function for both competing events. Below, we only consider
relapse (cause 1).

```{r, message=FALSE}
library(pseudo)

cutoffs <- c(5, 10, 15, 20, 25, 30, 35)

pseudo <- pseudoci(hd_train$time,
                   hd_train$status,
                   tmax = cutoffs)$pseudo$cause1     # we only consider cause 1
dim(pseudo)
```

`pseudo` contains the pseudo-values of the 692 patients (rows) at the 7 different
time points (columns) for relapse. 

The following code will merge the pseudo-values with our train dataset.

```{r}
merged_dataset <- NULL
for (j in seq_len(ncol(pseudo))) {
    merged_dataset <- rbind(merged_dataset,
                            cbind(hd_train,
                                  pseudo = pseudo[, j],
                                  tpseudo = cutoffs[j],
                                  id = seq_len(nrow(hd_train))))
}

merged_dataset <- merged_dataset[order(merged_dataset$id), ]
merged_dataset$tpseudo <- as.factor(merged_dataset$tpseudo)
```

We will now fit a generalised linear model using `geese()` function from R
package `geepack`. We set `mean.link = "cloglog"` to use a complementary
log-log link in order to recover Fine-Gray approach (alternative link functions
can also be used, e.g. @klein2008sas explored a logistic link function). 

```{r, message=FALSE}
library(geepack)

fit_pseudo <- geese(pseudo ~
                      tpseudo +
                      age +
                      sex +
                      trtgiven +
                      medwidsi +
                      extranod +
                      clinstg - 1,
                    data = merged_dataset,
                    id = id,
                    jack = TRUE,
                    scale.fix = TRUE,
                    family = gaussian,
                    mean.link = "cloglog",
                    corstr = "independence")
summary(fit_pseudo)
```

Covariate effects are shown below and are interpreted as logarithms of
sub-distribution hazard ratios. Because we used a complementary log-log link 
function, we observe that the estimated effects are akin to those obtained in 
the previous section. 

```{r}
exp(fit_pseudo$beta)
```

The estimated coefficients for the selected time points can be used to estimate 
the CIF, but this is only possible at the time-points used in the `cutoffs` grid. 

# Direct binomial

Binomial modelling approach can be fitted using the `timereg` R package.

To fit the model, we should first use the `Event()` function to obtain the
response variable to be used in the formula argument of the `comp.risk()`
function. The cause of interest must be explicitly defined through the cause
argument. The function supports several models. Here we focus on recovering the
Fine Gray approach.

```{r, warning=FALSE}
library(timereg)

fit_DB <- comp.risk(Event(time, status) ~
                      const(age) + # const for time-invariant
                      const(sex) +
                      const(trtgiven) + 
                      const(medwidsi) +
                      const(extranod) +
                      const(clinstg),
                    data = hd_train,
                    cause = 1,
                    n.sim = 5000,
                    model = "prop",       # Fine-Gray
                    resample.iid = 1,
                    # uses all covariates present for the censoring weights
                    cens.model = "cox")
summary(fit_DB)
```

According to the summary output;  age and treatment are significant. Below, we
show estimated hazard ratios.

```{r}
knitr::kable(round(exp(fit_DB$gamma), 4), col.names = c("Exp coef"))
```


In order to do predictions of the CIF, we use the `predict()` function. The
default makes predictions at all observed time points. Below we show estimated
incidence for subject 1 (red) and 3 (blue) in the tests set:

```{r fig.dim=c(6,4)}
par(mar = c(4, 4, 2, 0.1))
hd_test <- as.data.frame(hd_test[, 1:6])
pred_DB <- predict.timereg(fit_DB, newdata = hd_test[c(1, 3), 1:6], se = 0)
plot(pred_DB,
     multiple = 1,
     col = c("red", "blue"),
     ylim = c(0, 0.57),
     ylab = "Cumulative incidence")

```

In addition, one can also makes predictions at a specific time through the
argument `times`. Absolute risk predictions at time $t=30.5$ for the first 5
patients are shown:

```{r}
pred_DB.t <- predict.timereg(fit_DB, newdata = hd_test[1:5, 1:6], times = 30.5)
c(pred_DB.t$P1)
```

`timereg` permits fitting the model time varying effects. **warning problem**

```{r}
fit_DBtime <- comp.risk(Event(time, status) ~
                          age +
                          const(sex) +
                          const(trtgiven) + 
                          const(medwidsi) +
                          const(extranod) +
                          const(clinstg),
                        data = hd_train,
                        cause = 1,
                        model = "prop", # Fine-Gray
                        resample.iid = 1,
                        n.sim = 100,
                        # uses all covariates present for the censoring weights
                        cens.model = "cox")
summary(fit_DBtime)
```

# Dependent Dirichlet processes

> **_NOTE:_** `DPWeibull` is not available on CRAN because a dependency,
`binaryLogic`, was removed from CRAN as the maintainer's email address is no
longer accessible**

The formula definition uses `Hist()`.

An indication of a competing risk setting must be passed by setting
`comp = TRUE`.  Below, we run the MCMC chain for 5000 iterations, we discarded
3000 as part of the burn-in period, and to reduce autocorrelation, a thinning of
10 was used (default).


```{r, warning=FALSE , results='hide'}
library(DPWeibull)
library(prodlim) # for Hist() function
fit_DDP <- dpweib(Hist(time, status) ~
                    age +
                    sex +
                    trtgiven +
                    medwidsi +
                    extranod +
                    clinstg ,
                  data = hd_train,
                  comp = TRUE,
                  burnin = 1000,
                  iteration = 5000,
                  predtime = seq(from = min(hd_train$time),
                                 to = max(hd_train$time),
                  length =  40))
```

```{r}
summary(fit_DDP)
```

One can explore trace plots of the estimated parameters using the `coda`
library. This will help us to determine if the MCMC parameters employed when
fitting the model should be modified. For instance, below we show trace plots of
$\alpha$ and $\lambda$ for patient 1.

```{r, message=FALSE, fig.dim=c(5,2)}
library(coda)
par(mar = c(4, 4, 2, 0.1))
traceplot(as.mcmc(fit_DDP$alpharec1[, 1]))
traceplot(as.mcmc(fit_DDP$lambda0rec1[, 1]))
```

```{r}
(fit_DDP$betarec1[1:3, 1:3])
dim(fit_DDP$loghr.est)
fit_DDP$predtime
(fit_DDP$xmean)
fit_DDP$covnames
```

Below, we illustrate how to do predictions with `DPWeibull`. First, we show how
to estimate the CIF for two subjects in the test set at all observed event
times.

```{r fig.dim=c(6,4)}
par(mar = c(4, 4, 2, 0.1))
hd_test_subset <- as.data.frame(hd_test[c(1, 3), 1:6])

pred_DDP <- predict(fit_DDP, newdata = hd_test_subset)
```

Note that pred_DDP contains posterior samples for the 2 subjects at 40 observed
event times, as well as posterior estimated and credible intervals.

Below, we show the estimated CIF curves for patient 1 (red) and 3 (blue).

```{r fig.dim=c(6,4)}
plot(pred_DDP$tpred,
     pred_DDP$Fpred[2, ],
     type = "l",
     col = "red",
     xlab = "Time (years)",
     ylab = "Cumulative incidence")
lines(pred_DDP$tpred, pred_DDP$Fpred[1, ], col="blue")
plot(pred_DDP)
```

In addition, we can also make predictions at a specific time point, e.g.
$t=30.5$. Below, we show the estimated risk for the first 5 subjects in the test
data set:

```{r}
hd_test_subset <- as.data.frame(hd_test[1:5, 1:6])
pred_DDP.t <- predict(fit_DDP, newdata = hd_test_subset , tpred = 30.5)
pred_DDP.t$Fpred
```

# References

<div id="refs"></div>

# Session Info
```{r}
sessionInfo()
```
