---
title: "Data exploration"
author: "Karla Monterrubio-Gómez, Nathan Constantine-Cooke, and Catalina Vallejos"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: show
    toc: yes
    css: style.css
    theme: simplex
    highlight: textmate
    toc_float:
      collapsed: no
    includes:
      in_header: head.html
      before_body: navbar.html
    self_contained: false
  pdf_document:
    toc: yes
link-citations: yes
bibliography: references.bib
vignette: "%\\VignetteIndexEntry{Vignette Title} %\\VignetteEngine{knitr::rmarkdown}
  %\\VignetteEncoding{UTF-8}\n"
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  cache = FALSE
)
```

# Introduction

To demonstrate the usage of difference competing risks survival methods,
we use publicly available data.

The code below is used to load and pre-process the data. Additionally,
we randomly split the data into *training* and *testing* sets as that
will allow us to evaluate out-of-sample predictive performance when
comparing different approaches.

To avoid code duplication, a basic version of the code below is stored
as an R file which will be sourced at the start of each vignette.

# Data preparation

The dataset used here corresponds to the Hodgkin's disease (HD) study
described in Pintilie, 2006. The dataset comprises 865 patients
diagnosed with early stage (I or II) HD, and which were treated either
with radiation (RT) or with radiation and chemotherapy (CMT).

The recorded data includes:

-   age: Age (years)
-   sex: Sex, F=female and M=Male.
-   trtgiven: Treatment given, RT=Radiation, CMT=Chemotherapy and
    radiation
-   medwidsi: Size of mediastinum involvement, N=No, S=Small, L=Large
-   extranod: Extranodal disease, Y=Extranodal disease, N= Nodal disease
-   clinstg: Clinical stage, 1=Stage I, 2=Stage II
-   time: time to failure (years) calculated from the date of diagnosis
-   status: 0=censoring, 1=relapse and 2=death.

We now load and display the structure of the HD dataset:

```{r, message = FALSE, warning = FALSE}
require(readr)
hd <- data.frame(read_csv("../Data/HD/hd.csv",
                          col_types = cols(X1 = col_skip())))
str(hd)
```

To procede with the analysis, it is important to change the data type of
sex, trtgiven, medwidsi, and extranod from `character` to `factor`.
Similarly, we convert clinstg from `numeric` to `factor`.

```{r}
hd$sex      <- as.factor(hd$sex)
hd$trtgiven <- as.factor(hd$trtgiven)
hd$medwidsi <- as.factor(hd$medwidsi)
hd$extranod <- as.factor(hd$extranod)
hd$clinstg  <- as.factor(hd$clinstg)
```

# Data exploration

Now, we explore the number of events for each event type:

```{r, message=FALSE}
require(pander)
pander::pander(table(hd$status))
```

Thus, we have `r length(which(hd$status==0))` censored patients,
`r length(which(hd$status==1))` with relapse and
`r length(which(hd$status==2))` who died. From now on, we assume that
the event of interest is relapse, i.e. `status=1`.

In order to create a test set, we use stratified sampling to partition
our data in 80% for train and 20% for test.

```{r}
require(splitstackshape)
set.seed(2022)
split_data <- stratified(hd, c("status"), 0.8, bothSets = TRUE)
hd_train   <- split_data$SAMP1
hd_test    <- split_data$SAMP2
```

Now, we explore the number of observations per status in both train and
test set:

```{r}
pander::pander(table(hd_train$status))
```

```{r}
pander::pander(table(hd_test$status))
```

The following code is used to create the summary presented in Table 3:

```{r Table3}
library(table1)
table1(~ age + sex + trtgiven + medwidsi + extranod + clinstg | 
         factor(status), data = hd)
```

# References

::: {#refs}
:::

# Session Info

```{r}
sessionInfo()
```
