---
title: "Approaches based on a discrete survival times specification"
author: "Karla Monterrubio-Gómez, Nathan Constantine-Cooke, and Catalina Vallejos"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: show
    toc: yes
    css: style.css
    theme: simplex
    highlight: textmate
    toc_float:
      collapsed: no
link-citations: yes
bibliography: references.bib
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}

---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```



# Dataset

In order to demonstrate the methods, we employ publicly available data. 

The dataset used here corresponds to the Hodgkin’s disease (HD) study described
in Pintilie, 2006. The dataset comprises 865 patients diagnosed with early stage
(I or II) HD, and which were treated either with radiation (RT) or with
radiation and chemotherapy (CMT).

The recorded data includes:

* age: Age (years)
* sex: Sex, F=female and M=Male.
* trtgiven: Treatment given, RT=Radiation, CMT=Chemotherapy and radiation
* medwidsi: Size of mediastinum involvement, N=No, S=Small, L=Large
* extranod: Extranodal disease, Y=Extranodal disease, N= Nodal disease
* clinstg: Clinical stage, 1=Stage I, 2=Stage II
* time: time to failure (years) calculated from the date of diagnosis
* status: 0=censoring, 1=relapse and 2=death.

We now load and display the structure of the HD dataset:

```{r, message=FALSE, warning=FALSE}
library(readr)
hd <- data.frame(read_csv("../Data/HD/hd.csv",
                          col_types = cols(X1 = col_skip())))
str(hd)
```

To proceed with the analysis, it is important to change the data type of sex,
trtgiven, medwidsi, and extranod from character to factor. Similarly, we convert
clinstg from numeric to factor.

```{r}
hd$sex      <- as.factor(hd$sex)
hd$trtgiven <- as.factor(hd$trtgiven)
hd$medwidsi <- as.factor(hd$medwidsi)
hd$extranod <- as.factor(hd$extranod)
hd$clinstg  <- as.factor(hd$clinstg)
```

Now, we explore the number of events for each event type:

```{r}
table(hd$status)
```

Thus, we have 439 censored patients, 291 with relapse, and 135 who died. From
now on, we assume that the event of interest is relapse, i.e. status=1.

In order to create a test set, we use stratified sampling to partition our
dataset into 80% for train and 20% for test.

```{r, eval=FALSE}
library(splitstackshape)
set.seed(2022)
split_data <- stratified(hd, c("status"), 0.8, bothSets = TRUE)
hd_train   <- split_data$SAMP1
hd_test    <- split_data$SAMP2
```

Now, we explore the number of observations per status in both train and test
set:

```{r}
table(hd_train$status)
table(hd_test$status)
```

# BART
BART has a well documented vignette (@sparapani2021). Here, we focus only on demonstrating its usage for a CR setting, which corresponds to Section 5.3 in @sparapani2021. In the following, we fit the model with the two different likelihood formulations to compare the obtained estimates.
 
The first step is to recast the data creating dummy variables for all categorical covariates. This is done for both train and test sets:
```{r}
library(nnet)
library(survival)
library(BART)
library(stats)

xtrain = model.matrix(~. , hd_train[,c(1:6)])[,-1]
xtest = model.matrix(~. , hd_test[,c(1:6)])[,-1]
```


## Model formulation 1

The first method employs two binary likelihoods. The first one is a BART 
survival model for the time to the first event and the second model accounts for 
the probability of the event being of type $k=1$ given that the it occurred. The 
user can fit the model by using function `crisk2.bart()`. Note that the required 
binary event indicators $y_{ijk}$ can be constructed beforehand with 
`surv.pre.bart()` and passed to the function using the `y.train` argument. 
Instead, here we pass arguments `times` and `delta` which will construct event 
indicators internally. If we are interested in predictions, the test set can be 
passed directly when fitting the model through the argument `x.test`. Arguments 
to control the MCMC sampler are: `ndpost`, `nskip`, and `keepevery`. Their 
functionality is documented in the help file for the `crisk2.bart()` function. 


```{r, results='hide'}

bart1 <- crisk2.bart(x.train = xtrain, 
                     times = hd_train$time,   # needed if ytrain is not provided
                     delta = hd_train$status, # needed if ytrain is not provided
                     x.test = xtest, 
                     sparse = FALSE,          # set equal TRUE for variable selection 
                     type = 'pbart',
                     ntree = 30, numcut = 100, 
                     ndpost = 500, nskip = 100, keepevery = 10L, 
                     seed = 99)  
```
Note that the code shown above does not use multi-threading, but BART permits 
its usage by using `mc.crisk2.bart()` function instead of `crisk2.bart()`.

Studying convergence diagnosis of the MCMC chains is key to ensure the 
predictions are valid. In order to do that one needs to fit the model using the 
train set as test set, by setting the argument `x.test = xtrain`. Then we can 
use Geweke diagnostics to assess converge 

```{r, results='hide'}
bart1.chains <- crisk2.bart(x.train = xtrain, 
                     times = hd_train$time,   # needed if ytrain is not provided
                     delta = hd_train$status, # needed if ytrain is not provided
                     x.test = xtrain,         
                     sparse = FALSE,          # set equal TRUE for variable selection 
                     type = 'pbart',
                     ntree = 30, numcut = 100, 
                     ndpost = 500, nskip = 100, keepevery = 10L, 
                     seed = 99)  

h <- bart1.chains$K
samples <- bart1.chains$yhat.test[, 1:h]
z <- gewekediag(samples)$z
lim <- max(c(4, abs(z)))

# z scores vs. time for patient 1
plot(bart1.chains$times, z, ylim=c(-lim, lim), type='l',xlab='t', ylab='z')
lines(bart1.chains$times, rep(1.96, h), type='l', lty=2)
lines(bart1.chains$times, rep(-1.96, h), type='l', lty=2)
```

Notice that the Geweke statistics exceed the $95\%$ limits several times, 
suggesting the chains have not converged. Thus, the MCMC should be run for 
longer thining to obtain valid estimates. 

CIFs for the subjects in the test set can be obtained through `cif.test.mean()`. 
This provides the posterior mean across MCMC samples. In addition, credible 
intervals can be computed from the samples saved in `cif.test`. 
First, we re-organised the predicted CIF for cause 1 for the test dataset. The 
constructed matrix contains one row per subject and the columns correspond to 
the unique time points at which it is evaluated. Second, we compute $95\%$ 
credible intervals.

```{r}
cif.pred <- matrix(bart1$cif.test.mean, nrow=nrow(xtest), byrow = TRUE )

# Compute 95% credible intervals and put in matrix format:
cif.025 <- apply(bart1$cif.test, 2, quantile, probs = 0.025) 
cif.025 <- matrix(cif.025, nrow=nrow(xtest), byrow = TRUE)
cif.975 <- apply(bart1$cif.test, 2, quantile, probs = 0.975) 
cif.975 <- matrix(cif.975, nrow=nrow(xtest), byrow = TRUE)
```

We show CIF curves for the first (red) and second (blue) individuals in the test 
set along with its corresponding credible intervals:

```{r, fig.dim=c(6,4)}
par(mar = c(4, 4, 2, 0.1))

plot(bart1$times,
     cif.pred[1,],
     type = "l",
     col = "red",
     ylim = c(0, 0.9),
     xlab = "Time (years)",
     ylab = "Cumulative incidence")
points(bart1$times,cif.025[1,], col = "red", type ='s', lwd = 1, lty = 2)
points(bart1$times,cif.975[1,], col = "red", type = 's', lwd = 1, lty = 2)
lines(bart1$times,  cif.pred[2,], col="blue")
points(bart1$times,cif.025[2,], col = "blue", type ='s', lwd = 1, lty = 2)
points(bart1$times,cif.975[2,], col = "blue", type = 's', lwd = 1, lty = 2)
```

Similar to other approaches, BART permits to do predictions of the CIF at
a specific time point (e.g. $t=5$ years) for a new dataset (e.g. `hd_test`). 
Note that predictions are only provided for time-points present in the training
dataset. Below, we show results for the first 5 subjects in the test set.

```{r}
BART.pred <- matrix(bart1$cif.test.mean, nrow=nrow(xtest), byrow = TRUE )
BART.pred[1:5,which(bart1$times == 5)]
```

Once the model was fitted and we have confirmed convergence of the MCMC chains, 
we can do predictions by making a call to the `predict.crisk2bart()` function. 
For instance:

```{r}
pre <- surv.pre.bart(x.train=xtrain, x.test=xtest, 
                     times=hd_train$time, delta=hd_train$status)

bart1.pred <- mc.crisk2.pwbart(pre$tx.test, pre$tx.test,
                         bart1.chains$treedraws, bart1.chains$treedraws2)

pre <- surv.pre.bart(x.train=xtrain, x.test=xtest, times=hd_train$time, 
                     delta =hd_train$status)

bart1.pred <- predict(bart1.chains, newdata=pre$tx.test, newdata2=pre$tx.test)

cif.pred <- matrix(bart1.pred$cif.test.mean, nrow=nrow(xtest), byrow = TRUE )
cif.025 <- apply(bart1.pred$cif.test, 2, quantile, probs = 0.025) 
cif.025 <- matrix(cif.025, nrow=nrow(xtest), byrow = TRUE)
cif.975 <- apply(bart1.pred$cif.test, 2, quantile, probs = 0.975) 
cif.975 <- matrix(cif.975, nrow=nrow(xtest), byrow = TRUE)

plot(bart1.chains$times,
     cif.pred[1,],
     type = "l",
     col = "red",
     ylim = c(0, 1),
     xlab = "Time (years)",
     ylab = "Cumulative incidence")
points(bart1.chains$times,cif.025[1,], col = "red", type ='s', lwd = 1, lty = 2)
points(bart1.chains$times,cif.975[1,], col = "red", type = 's', lwd = 1, lty = 2)
lines(bart1.chains$times,  cif.pred[3,], col="blue")
points(bart1.chains$times,cif.025[3,], col = "blue", type ='s', lwd = 1, lty = 2)
points(bart1.chains$times,cif.975[3,], col = "blue", type = 's', lwd = 1, lty = 2)

```

### DART

It is possible to employ a sparse Dirichlet prior for variable selection 
(DART model). This will help us to determine variable importance. In order to 
fit such model we use again `crisk2.bart()` function and set the `sparse` 
argument equal to `TRUE`.

```{r, results='hide'}
dart1 <- crisk2.bart(x.train = xtrain, 
                     times = hd_train$time,   # needed if ytrain is not provided
                     delta = hd_train$status, # needed if ytrain is not provided
                     x.test = xtest, 
                     sparse = TRUE,          # set equal TRUE for variable selection 
                     type = 'pbart',
                     ntree = 30, numcut = 100, 
                     ndpost = 500, nskip = 200, keepevery = 10L, 
                     seed = 99)  
```

The output of the function is the same as discussed in the previous section. Therefore,  CIFs and survival functions can be estimated as shown before.

A plot that illustrates which variables to select is shown:
```{r, fig.dim=c(6,4)}
P = ncol(xtrain)    # use to set thereshold probability for each covariate

dart1$varprob.mean > 1/P

plot(dart1$varprob.mean[-1], 
     ylab='Selection Probability', 
     ylim=c(0, 1), 
     pch = 1+45*(dart1$varprob.mean <= 1/P))
lines(c(1, P+1), c(1/P, 1/P))
```

According to the plot above only age and treatment are relevant.


## Model formulation 2

This approach is discussed in Section 3.2 of @Bart2020 and fits also two separate BART probit models. The first model, corresponds to the conditional probability of a cause $k=1$ event at a given time. The second, models the conditional probability of an event of type $k=2$ at a specific time, given that the individual is still at risk and did not experience a type $k=1$ event.
In this case, the model is fit with function `crisk.bart`:

```{r, results='hide'}

bart2 <- crisk.bart(x.train = xtrain, times = hd_train$time, delta=hd_train$status,
           x.test = xtest, 
           sparse=FALSE, 
           type='pbart',
           ntree = 30, numcut = 100, 
           ndpost = 500, nskip = 100, keepevery = 10L, 
           seed=99)
# Parallel computation of the model is available using mc.crisk.bart
```

The output is the same as in model formulation 1. 


```{r}
cif2.pred <- matrix(bart2$cif.test.mean, nrow=nrow(xtest), byrow = TRUE )

# Compute 95% credible intervals and put in matrix format:
cif2.025 <- apply(bart2$cif.test, 2, quantile, probs = 0.025) 
cif2.025 <- matrix(cif2.025, nrow=nrow(xtest), byrow = TRUE)
cif2.975 <- apply(bart2$cif.test, 2, quantile, probs = 0.975) 
cif2.975 <- matrix(cif2.975, nrow=nrow(xtest), byrow = TRUE)
```

We show CIF curves for patient 1 (red) and 3 (blue) in the test set along with its corresponding credible intervals:
```{r, fig.dim=c(6,4)}
par(mar = c(4, 4, 2, 0.1))

plot(bart2$times,
     cif2.pred[1,],
     type = "l",
     col = "red",
     ylim = c(0, 1),
     xlab = "Time (years)",
     ylab = "Cumulative incidence")
points(bart2$times,cif2.025[1,], col = "red", type ='s', lwd = 1, lty = 2)
points(bart2$times,cif2.975[1,], col = "red", type = 's', lwd = 1, lty = 2)
lines(bart2$times,  cif2.pred[3,], col="blue")
points(bart2$times,cif2.025[3,], col = "blue", type ='s', lwd = 1, lty = 2)
points(bart2$times,cif2.975[3,], col = "blue", type = 's', lwd = 1, lty = 2)
```


```{r}
surv2.pred <- matrix(bart2$surv.test.mean, nrow=nrow(xtest), byrow = TRUE )
surv2.pred[1:5,which(bart2$times == 30.5)]

# Compute 95% credible intervals and put in matrix format:
surv2.025 <- apply(bart2$surv.test, 2, quantile, probs = 0.025) 
surv2.025 <- matrix(surv2.025, nrow=nrow(xtest), byrow = TRUE)
surv2.975 <- apply(bart2$surv.test, 2, quantile, probs = 0.975) 
surv2.975 <- matrix(surv2.975, nrow=nrow(xtest), byrow = TRUE)
```

A plot for the first subject is shown below:
```{r,  fig.dim=c(6,4)}

plot(bart2$times,
     surv2.pred[1,],
     type = "l",
     col = "red",
     ylim = c(0, 1),
     xlab = "Time (years)",
     ylab = "Survival")
points(bart2$times,surv2.025[1,], col = "red", type ='s', lwd = 1, lty = 2)
points(bart2$times,surv2.975[1,], col = "red", type = 's', lwd = 1, lty = 2)

```


Note that the estimates of the 2 model formulations differ. The next plots compare CIFs for patient 1, under the 2 different models to show such differences:
```{r}

plot(bart1$times,bart1$cif.pred[1,],lwd=2,type="l", col=2, ylim=c(0,1), 
     main="Comparison of different formulations for test patient 1",
     xlab="Time", ylab="CIF(t)")
lines(bart2$cif2.pred[2,], col=3,lwd=3)
      legend("topright", 
       legend=c("Formulation 1", "Formulation 2"),
       col=c("#009999", "#FFCC00"), lty=c(1,1))
```

# References

<div id="refs"></div>

# Session Info
```{r}
sessionInfo()
```
