---
title: "Other methods"
author: "Karla Monterrubio-Gómez, Nathan Constantine-Cooke, and Catalina Vallejos"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: show
    toc: yes
    css: style.css
    theme: simplex
    highlight: textmate
    toc_float:
      collapsed: no
link-citations: yes
bibliography: references.bib
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}

---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```



# Dataset

In order to demonstrate the methods, we employ publicly available data. 

The dataset used here corresponds to the Hodgkin’s disease (HD) study described
in @pintilie2006competing. The dataset comprises 865 patients diagnosed with early stage
(I or II) HD, and which were treated either with radiation (RT) or with
radiation and chemotherapy (CMT).

The recorded data includes:

* age: Age (years)
* sex: Sex, F=female and M=Male.
* trtgiven: Treatment given, RT=Radiation, CMT=Chemotherapy and radiation
* medwidsi: Size of mediastinum involvement, N=No, S=Small, L=Large
* extranod: Extranodal disease, Y=Extranodal disease, N= Nodal disease
* clinstg: Clinical stage, 1=Stage I, 2=Stage II
* time: time to failure (years) calculated from the date of diagnosis
* status: 0=censoring, 1=relapse and 2=death.

We now load and display the structure of the HD dataset:

```{r, message=FALSE, warning=FALSE}
library(readr)
hd <- data.frame(read_csv("../Data/HD/hd.csv",
                          col_types = cols(X1 = col_skip())))
str(hd)
```

To proceed with the analysis, it is important to change the data type of sex,
trtgiven, medwidsi, and extranod from character to factor. Similarly, we convert
clinstg from numeric to factor.

```{r}
hd$sex      <- as.factor(hd$sex)
hd$trtgiven <- as.factor(hd$trtgiven)
hd$medwidsi <- as.factor(hd$medwidsi)
hd$extranod <- as.factor(hd$extranod)
hd$clinstg  <- as.factor(hd$clinstg)
```

Now, we explore the number of events for each event type:

```{r}
table(hd$status)
```

Thus, we have 439 censored patients, 291 with relapse, and 135 who died. From
now on, we assume that the event of interest is relapse, i.e. status=1.

In order to create a test set, we use stratified sampling to partition our
dataset into 80% for train and 20% for test.

```{r, eval=FALSE}
library(splitstackshape)
split_data <- stratified(hd, c("status"), 0.8, bothSets = TRUE)
hd_train   <- split_data$SAMP1
hd_test    <- split_data$SAMP2
```


```{r, echo=FALSE}
hd_train <- read_csv("../Data/HD/hd_train.csv", show_col_types = FALSE)[, -1]
hd_test <- read_csv("../Data/HD/hd_test.csv", show_col_types = FALSE)[, -1]

hd_train$sex      <- as.factor(hd_train$sex)
hd_train$trtgiven <- as.factor(hd_train$trtgiven)
hd_train$medwidsi <- as.factor(hd_train$medwidsi)
hd_train$extranod <- as.factor(hd_train$extranod)
hd_train$clinstg  <- as.factor(hd_train$clinstg)

hd_test$sex      <- as.factor(hd_test$sex)
hd_test$trtgiven <- as.factor(hd_test$trtgiven)
hd_test$medwidsi <- as.factor(hd_test$medwidsi)
hd_test$extranod <- as.factor(hd_test$extranod)
hd_test$clinstg  <- as.factor(hd_test$clinstg)
```

Now, we explore the number of observations per status in both train and test
set:

```{r}
table(hd_train$status)
table(hd_test$status)
```



# Vertical modelling

```{r}
library(splines)
library(survival)
```

# Random survival forests

The authors provide an online vignette for RSF [here](https://www.randomforestsrc.org/articles/competing.html) (@HemantCompeting).

The argument `splitrule` is key when determining the type of analysis we are interested on. When the objective is to do a cause-specific hazard analysis we must set `splitrule="logrank"` and the argument `cause` should indicate the event type of interest. In contrast, if one is interested on long term predictions, `splitrule="logrankCR"` is more appropiate, as it considers the CIF.

Below, we fit both models:

## Generalized log-rank splitting rule (logrank)

In order to assess variable importance, we set argument `importance=TRUE`
by.root which bootstraps the data by sampling
with or without replacement (without replacement is the default;
```{r}
library(randomForestSRC)

rsf.lr <- rfsrc(Surv(time, status) ~ ., 
                data = hd_train,
                ntree = 100,
                splitrule = "logrank", 
                cause = c(1,0),         # to indicate that first event is of interest
                bootstrap = "by.root",
                importance = TRUE,
                save.memory= TRUE,     # useful for big datasets
                seed = 100
                )
plot.competing.risk(rsf.lr)

```

Note that if missing values are present, it is possible to imputem them by setting the argument `na.action=na.impute`.

```{r}
csh.oob <- chf.oob 
cif.oob <- rsf.lr$cif.oob
```

Predictions
```{r}
test.lr <- predict(rsf.lr, newdata = dh_test,
                   importance = TRUE, csv = TRUE)
```

## Modified Gray’s test splitting rule (logrankCR)

```{r}

rsf.lrCR <- rfsrc(Surv(time, status) ~ ., 
                data = hd_train,
                ntree = 100,
                splitrule = "logrankCR", 
                cause = c(1,0),         # to indicate that first event is of interest
                importance = TRUE,
                save.memory= TRUE,     # useful for big datasets
                seed = 100
                )
plot.competing.risk(rsf.lrCR)

```

Note that if missing values are present, it is possible to imputem them by setting the argument `na.action=na.impute`.

```{r}
csh.oob <- rsf.lrCR$chf.oob 
cif.oob <- rsf.lrCR$cif.oob
```

Predictions
```{r}
test.lrCR <- predict(rsf.lrCR, newdata = dh_test,
                   importance = TRUE, csv = TRUE)
```


# References
<div id="refs"></div>

# Session Info
```{r}
sessionInfo()
```