---
title: "Comparison of predictions"
author: "Karla Monterrubio-Gómez, Nathan Constantine-Cooke, and Catalina Vallejos"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: show
    toc: yes
    css: style.css
    theme: simplex
    highlight: textmate
    includes:
      in_header: head.html
      before_body: navbar.html
    self_contained: false
    toc_float:
      collapsed: no
link-citations: yes
bibliography: references.bib
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  cache = FALSE
)
```

# Introduction

The dataset used here corresponds to the Hodgkin’s disease (HD) study described
in Pintilie, 2006. The data was loaded and pre-processed using the script that
is sourced below. The latter creates three data objects:

- `hd`: all the samples in the original data
- `hd_train`: a random sample of the data to be used as a training set
- `hd_test`: a random sample of the data to be used as a testing set

Details about data pre-processing steps can be found [here](https://github.com/VallejosGroup/CompRisksVignettes/blob/main/Source/DataPreparation.Rmd)

```{r}
source("../Aux/data_prep.R")
```

This document compares predictions generated the different approaches for individuals in the test set, focusing on the risk of the primary event type
(relapse) at $t=5$ years. 

# Comparing predictions at $t=5$

## Loading the predictions

Here, we load the predicted probabilities of observing an event of type 1 by
$t=5$ that were obtained by different methods

```{r load_pred}
if (file.exists("/.dockerenv")){
  pred_CS <- read.csv("/Outputs/pred_CS.csv")
  pred_CIF <- read.csv("/Outputs/pred_CIF.csv")
  pred_Others <- read.csv("/Outputs/pred_Others.csv")
  pred_BART <- read.csv("/Outputs/pred_BART.csv")
} else {
  pred_CS <- read.csv("../Outputs/pred_CS.csv")
  pred_CIF <- read.csv("../Outputs/pred_CIF.csv")
  pred_Others <- read.csv("../Outputs/pred_Others.csv")
  pred_BART <- read.csv("../Outputs/pred_BART.csv")
}

pred_all <- merge(pred_CS, pred_CIF, by = "testID")
pred_all <- merge(pred_all, pred_Others, by = "testID")
pred_all <- merge(pred_all, pred_BART, by = "testID")

names(pred_all)
```

The data above includes predictions obtained using the following methods:

* `coxph_riskRegression`: Cause-specific Cox PH (CS Cox) model 
* `coxphPL_riskRegression`: Cause-specific Cox PH (CS Cox) model using a product
limit (PL) approach to calculated predicted probabilities. The latter ensures
that (in the limit) the event-specific probabilities sum up to 1 across events. 
* `FG_riskRegression`: Fine & Gray (F&G) regression   
* `DirectBinomial`: Direct binomial        
* `DPWeibull`: Dependent Dirichlet Process        
* `RF_logrankCR`: Random Forests (RF) using a modified Gray's criteria as the 
splitting rule, focusing on the first event type. The latter aims to identify 
covariates that affect the CIF for the first event type.
* `RF_logrankCR.av`: as above, but the splitting rule considers the average 
between the first and second event types.
* `crisk2.bart`: BART using its first formulation    
* `crisk2.bart_dart`: DART - a variation of BART (first formulation) that
enables feature selection
* `crisk.bart`: BART using its second formulation    

## Pairwise comparison

As shown below, all methods led to correlated predictions. The greatest similarity was between the classical approaches (CS Cox, F&G, Direct binomial).
BART, under its second formulation, also led to similar predictions. RF 
exhibited the greatest discrepancy with respect to such methods above.
Note that for the different BART approaches and DPWeibull, predicted probabilities may be affected by lack of convergence of the algorithm. 

The following code was used to create Figure 2:

```{r, fig.width=12, fig.height=12}
library(GGally)

# Code to control axis limits was adapted from: https://stackoverflow.com/questions/53277656/how-to-define-facet-axis-limits-in-ggpairs-function
scatter_limitRange <- function(data, mapping, ...) { 
  ggplot(data = data, mapping = mapping, ...) + 
    geom_point(..., col = "gray", alpha = 0.5) + 
    geom_abline(intercept = 0, slope = 1) +
    scale_x_continuous(limits = c(min(pred_all[,-1]), max(pred_all[,-1]))) +
    scale_y_continuous(limits = c(min(pred_all[,-1]), max(pred_all[,-1]))) +
    theme_classic()
}
dens_limitRange <- function(data, mapping, ...) {
  ggplot(data = data, mapping = mapping, ...) + 
    geom_density(..., col = "gray", fill = "gray") + 
    scale_x_continuous(limits = c(min(pred_all[,-1]), max(pred_all[,-1]))) +
    theme_classic()
}

my.labels <- c("CS Cox", "CS Cox (PL)", "F&G", "Direct Binomial", 
               "DPWeibull", "RF logrankCR (1)", "RF logrankCR (av)", 
               "BART (1)", "DART (1)", "BART (2)")

ggpairs(pred_all, columns = 2:11, progress = FALSE,
        lower = list(continuous = scatter_limitRange),
        diag = list(continuous = dens_limitRange),
        columnLabels = my.labels)

if (file.exists("/.dockerenv")){ # running in docker
  ggsave("/Outputs/Comparison_predictions_t5.pdf", device = "pdf")
} else {
  ggsave("../Outputs/Comparison_predictions_t5.pdf", device = "pdf")
}
```

To evaluate the predictive performance for each methods, users can follow the
examples provided by [van Geloven et al](https://github.com/survival-lumc/ValidationCompRisks).

## Predictions versus covariates

```{r, fig.width=16, fig.height=5}
# combine test dataset with the predictions
pred_all <- cbind(pred_all, hd_test)

library(tidyverse)
library(patchwork)

y.min <- with(pred_all, min(FG_riskRegression, DPWeibull, RF_logrankCR))
y.max <- with(pred_all, max(FG_riskRegression, DPWeibull, RF_logrankCR))

p_FG <- ggplot(data = pred_all, 
                aes(x = age, y = FG_riskRegression, 
                    col = sex, pch = trtgiven)) +
  geom_point(cex = 2) + theme_classic() +
  ylab("Prob relapse before 5 years (Fine & Gray)") +
  ylim(y.min, y.max) +
  scale_color_manual(values = c("#B3B6B7", "#000000"))

p_DW <- ggplot(data = pred_all, 
                aes(x = age, y = DPWeibull, 
                    col = sex, pch = trtgiven)) +
  geom_point(cex = 2) + theme_classic() +
  ylab("Prob relapse before 5 years (DPWeibull)") +
  ylim(y.min, y.max) +
  scale_color_manual(values = c("#B3B6B7", "#000000"))

p_RF <- ggplot(data = pred_all, 
                aes(x = age, y = RF_logrankCR, 
                    col = sex, pch = trtgiven)) +
  geom_point(cex = 2) + theme_classic() +
  ylab("Prob relapse before 5 years (RF; logrankCR)") +
  ylim(y.min, y.max) +
  scale_color_manual(values = c("#B3B6B7", "#000000"))

p_FG + p_DW + p_RF + plot_annotation(tag_levels = 'A') + 
  plot_layout(guides = 'collect')

if (file.exists("/.dockerenv")){ # running in docker
  ggsave("/Outputs/Comparison_predictions_t5_covariates.pdf", device = "pdf")
} else {
  ggsave("../Outputs/Comparison_predictions_t5_covariates.pdf", device = "pdf")
}
```

# References
<div id="refs"></div>

# Session Info
```{r}
sessionInfo()
```
