---
title: "Comparison of predictions"
author: "Karla Monterrubio-Gómez, Nathan Constantine-Cooke, and Catalina Vallejos"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: show
    toc: yes
    css: style.css
    theme: simplex
    highlight: textmate
    includes:
      in_header: head.html
      before_body: navbar.html
    self_contained: false
    toc_float:
      collapsed: no
link-citations: yes
bibliography: references.bib
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  cache = TRUE
)
```


# Loading predictions

Here, we load the predicted probabilities of observing an event of type 1 by
$t=5$ that were obtained by different methods


```{r load_pred}
if (file.exists("/.dockerenv")){
  pred_CIF <- read.csv("/Predictions/pred_CIF.csv")
  pred_Others <- read.csv("/Predictions/pred_Others.csv")
  pred_BART <- read.csv("/Predictions/pred_BART.csv")
} else {
  pred_CIF <- read.csv("../Predictions/pred_CIF.csv")
  pred_Others <- read.csv("../Predictions/pred_Others.csv")
  pred_BART <- read.csv("../Predictions/pred_BART.csv")
}

pred_All <- merge(pred_CIF, pred_Others, by = "testID")
pred_All <- merge(pred_All, pred_BART, by = "testID")
```

# Comparing predictions

As shown below, all methods led to correlated predictions. As expected we 
observe greater similarity between methods based on the CIF specification
(Fine and Gray regression, direct binomial and DPWeibul). The three BART 
variations also led to similar predictions (discrepancies may be due to lack
of convergence). 

```{r cor_pred, fig.height=8, fig.width=8}
plot(pred_All[,-1], xlim = c(0, 0.65), ylim = c(0, 0.65))
```

To evaluate the predictive performance for each methods, users can follow the
examples provided by [van Geloven et al](https://github.com/survival-lumc/ValidationCompRisks).

# References
<div id="refs"></div>

# Session Info
```{r}
sessionInfo()
```
